---
title: ""
execute:
    echo: False
format: 
  html:
    self-contained: true
    toc: true
    theme: superhero
    title-block-banner: "194c66"
    author: "Julianna Mack & Ben Hopkins" 
    date: 3/17/2025
    date-format: short
editor: 
  markdown: 
    wrap: 72
---

::: header
<img src="../../images/cbo_logo_white_desktop.svg"/>
:::

# Premium Growth Model Annual Report: 2025

##### Last Updated: `r format(Sys.Date(), "%A %b %d, %Y")`

##### Author: Julianna Mack & Ben Hopkins

```{r import_packages, warning=FALSE, message=FALSE, results='hide'}
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(ggplot2)
library(patchwork)

# Parameters
baseline_vintage <- 2025
baseline_vintage_prev <- baseline_vintage - 1
projection_start <- as.numeric("2024")  # Start year for current baseline
projection_start_prev <- as.numeric("2024")  # Start year for previous baseline
projection_end <- as.numeric("2035")  # End year for current baseline
projection_end_prev <- as.numeric("2034")  # End year for previous 

# Define the head directory of the repository, which is three levels above the directory where this program is saved
dir_repo <- normalizePath(file.path(getwd(), "../../.."))

# Define output and nongroup baseline directories
dir_output <- paste0(dir_repo, "/output/baseline_", baseline_vintage, "/")
dir_nongroup_baseline <- paste0(dir_repo, "/prepped_data/nongroup_baseline/", baseline_vintage_prev, "/")

#  Define input files
file_pgm_output <- paste0("pgm_", baseline_vintage, ".csv")
file_previous_ng_baseline <- paste0(baseline_vintage_prev, "_baseline_nongroup.csv")

#  Load final output file from PGM
pgm_output <- read_csv(paste0(dir_output, file_pgm_output)) 

# Define series names dynamically
prem_demo_prev <- paste0("prem_demo_b", baseline_vintage_prev)
prem_demo <- paste0("prem_demo_b", baseline_vintage)
pchange_prem_demo_prev <- paste0("pchange_", prem_demo_prev)
pchange_prem_demo <- paste0("pchange_", prem_demo)
prem_prev <- paste0("prem_b", baseline_vintage_prev)
prem <- paste0("prem_b", baseline_vintage)
pchange_prem_prev <- paste0("pchange_", prem_prev)
pchange_prem <- paste0("pchange_", prem)
level_prev <- prem_demo_prev
level_step1 <- paste0(prem_demo_prev, "_step1")
level_step2 <- paste0(prem_demo_prev, "_step2")
level_step3 <- paste0(prem_demo_prev, "_step3")
level_step4 <- paste0(prem_demo_prev, "_step4")
level_current <- prem_demo
level_jct_prev <- prem_prev
level_jct_current <- prem
pchange_prev <- pchange_prem_demo_prev
pchange_step1 <- paste0(pchange_prem_demo_prev, "_step1")
pchange_step2 <- paste0(pchange_prem_demo_prev, "_step2")
pchange_step3 <- paste0(pchange_prem_demo_prev, "_step3")
pchange_step4 <- paste0(pchange_prem_demo_prev, "_step4")
pchange_current <- pchange_prem_demo
pchange_jct_prev <- pchange_prem_prev
pchange_jct_current <- pchange_prem

# Format the data for plotting
df_long <- pgm_output %>%
    pivot_longer(
    cols = starts_with("cy"),            # Columns with years (cy1987, cy1988, ...)
    names_to = "year",                   # Name for the new 'year' column
    values_to = "value"                  # Name for the new 'value' column
  ) %>%
  mutate(year = as.numeric(sub("cy", "", year)), .before = "value") %>%
  select(-first_historical_year, -specification)

# Define the color, shape, and label values with dynamic names
series_vector <- c(
    level_prev, 
    level_step1,
    level_step2,
    level_step3,
    level_step4,
    level_current,
    level_jct_prev, 
    level_jct_current,
    pchange_prev, 
    pchange_step1,
    pchange_step2,
    pchange_step3,
    pchange_step4,
    pchange_current,
    pchange_jct_prev, 
    pchange_jct_current
  )
  
color_values <- setNames(
  rep(
      c(
        "#EE1818", 
        "#A23616",
        "#565514",
        "#0A7413",
        "#2B607E",
        "#4C4CE9",
        "#EE1818", 
        "#4C4CE9"
      ), 
    2)
  ,
  series_vector
)

shape_values <- setNames(
  rep(
      c(
        16, # Filled circle
        17, # Filled triangle
        18, # Filled diamond
        4,  # X
        15, # Square
        3,  # +
        16, # Filled circle
        15 # Square
      ),
    2),
  series_vector
)

label_values <- setNames(
  rep(
      c(
        paste0(baseline_vintage_prev, " baseline"), 
        "Updated Historicals",
        "Updated Macroeconomic Forecast",
        "Updated Coefficients",
        "Updated Specification",
        paste0(baseline_vintage, " baseline"),
        paste0(baseline_vintage_prev, " baseline"), 
        paste0(baseline_vintage, " baseline")
      ),
    2),
  series_vector
)

forecast_plots <- function(level_series, pchange_series, combined_title) {
  # Declare function to create the main plots comparing premium projections
  #' Generate Projection Plots
  #'
  #' Creates combined level and percent change projection plots based on provided series data.
  #'
  #' @param level_series A character vector specifying the series to be plotted as levels.
  #' @param pchange_series A character vector specifying the series to be plotted as percent changes.
  #' @param combined_title A string representing the title for the combined plot.
  #'
  #' @return A combined ggplot object displaying both level and percent change projections.

  subplot <- function(subplot_series, is_levels = TRUE) { 
    #' Generate a Subplot for Specified Data Series
    #'
    #' This function creates a ggplot-based subplot for the given data series. It differentiates between historical and projection data by styling historical data as solid lines and projection data as dashed lines. The y-axis can be formatted to display either levels in dollars or percent changes based on the `is_levels` parameter.
    #'
    #' @param subplot_series A character vector specifying the series to include in the subplot.
    #' @param is_levels A logical value indicating whether to format the y-axis as levels (`TRUE`) or percent changes (`FALSE`). Defaults to `TRUE`.
    #'
    #' @return A ggplot object representing the generated subplot.

    # Duplicate the dataset, with the series truncated at the last historical year in one copy
    # This allows us to plot the historical data as a solid line and the projection data as a dashed line
    plot_data_full <- df_long %>%
      filter(
        year >= 1998,
        series %in% subplot_series 
      ) |>
      mutate(actual = "Projection")
    plot_data_actual <- plot_data_full %>%
      filter(year <= last_historical_year) |>
      mutate(actual = "Historical")
    plot_data <- bind_rows(
      plot_data_actual,
      plot_data_full
    )

    # Define the plot
    return_plot <- plot_data |>
      ggplot(aes(x = year, y = value, color = series, shape = series, linetype = actual)) +
      geom_hline(yintercept = 0, color = "grey") +
      geom_vline(xintercept = projection_start, color = "grey", linetype = "dashed") + 
      geom_line(size = 1.2) +
      geom_point(size = 2.5) +
      scale_linetype_manual(values = c("Projection" = "dashed", "Historical" = "solid")) + 
      scale_color_manual(
        values = color_values,
        labels = label_values
      ) + 
      scale_shape_manual(
        values = shape_values,
        labels = label_values
      ) + 
      labs(
        x = NULL,
        color = NULL,
        shape = NULL
      ) +
      guides(linetype = "none")

      # Format the y-axis based on whether the series are levels or percent changes
      if (is_levels) {
        return_plot <- return_plot + 
          # Format the y-axis labels as dollars and start the y-axis at 0
          scale_y_continuous(labels = scales::dollar, limits = c(0, NA)) +
          labs(y = "Dollars")
      }
      else {
        return_plot <- return_plot + 
          # Format the y-axis labels as percentages and start the y-axis at 0
          scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
          labs(y = "Percent Change")
      }
    return(return_plot)
  }

  # Generate the plots for the level and percent change series
  level_plot <- subplot(level_series, is_levels = TRUE)
  growth_plot <- subplot(pchange_series, is_levels = FALSE)

  # Generate and return the combined plot
  combined_plot <- (level_plot | growth_plot) +
    plot_layout(guides = "collect") &
    plot_annotation(
      title = combined_title
    ) &
    theme(
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
        legend.position = "top",
        legend.title = element_blank(),
        # Set the background of the markers in the legend to be transparent
        legend.key = element_rect(fill = "transparent")
    )
  return(combined_plot)
}
```

```{r calc_increase, warning=FALSE, message=FALSE, results='hide'}
# Calculate the increase in private health expenditures per capita in 2034 
# between prem_demo_b2024 and prem_demo_b2025
increase <- pgm_output |>
    select(series, cy2034) |>
    filter(series %in% c("prem_demo_b2024", "prem_demo_b2025")) |>
    spread(series, cy2034) |>
    mutate(
        pct_increase = (prem_demo_b2025 - prem_demo_b2024) / prem_demo_b2024,
        dollar_increase = prem_demo_b2025 - prem_demo_b2024
      )

pct_increase <- increase |>
    pull(pct_increase) |>
    # Format as a percentage
    scales::percent(accuracy = 0.1)

dollar_increase <- scales::label_dollar(accuracy=1)(increase |> pull(dollar_increase))
```

## Introduction

The Congressional Budget Office uses the premium growth model (PGM) to
project health expenditures per privately insured person through the
10-year baseline period. The PGM generates factors used in CBO's
health insurance simulation model, HISIM2, to project increases in health spending, employment-based insurance
(EBI) premiums, and employer contributions to health
reimbursement accounts (HRAs) and health savings accounts (HSAs). The
factors are also used to project growth in plan characteristics such as maximum
out-of-pocket spending for EBI plans and for nongroup plans, and deductibles
for nongroup plans. Additionally, the output of the PGM is used in HISIM2 to project growth in premiums for nongroup plans. The PGM's output is also used by the Joint Committee on
Taxation and is used for other purposes by CBO.

The PGM is updated by CBO each year, and those updates are incorporated
into CBO's spring baselines. The most recent PGM projections are for the
spring 2025 baseline, and the previous projections were for the spring
2024 baseline. One way to compare the two sets of projections is to
focus on the projected expenditures for health care services per capita
by insurers in the last year they share in common (2034). Looking at the
figure below, the 2025 updates to the PGM have increased projected
expenditures per capita in 2034 by about `r dollar_increase`
(`r pct_increase`). CBO now projects higher per capita expenditures than
it did last year after incorporating newly available historical data
indicating rapid premium growth in 2022 and 2023 and consulting with
stakeholders, who suggested that high growth rates are likely to persist
through at least 2025. The surge in expenditures among the privately
insured partly reflects the adoption of costly pharmaceuticals, most
notably GLP-1s, but it also reflects broader growth in spending on
hospital care and mental health care. CBO has also significantly updated
the specification of the regression model. Those technical updates
have smoothed the path of projected premium growth but did not
significantly alter the projected level of expenditures at the end of
the period.

```{r compare_forecast, warning=FALSE, message=FALSE, results='hide'}

level_series_compare <- c(level_prev, level_current)
pchange_series_compare <- c(pchange_prev, pchange_current)
combined_plot_compare <- forecast_plots(level_series_compare, pchange_series_compare, "Previous and Current Projections")
combined_plot_compare
```

## Overview of the Premium Growth Model (PGM) and Updates for the 2025 Baseline

The PGM is a set of two time-series regression models used to project
growth in health expenditures per privately insured person; those
projections serve as the basis for premiums in CBO's modeling. Health
expenditures, in this context, include amounts paid by the health
insurer and administrative expenses incurred by the insurer and exclude
cost sharing paid by the patient at the point of service and
out-of-pocket spending on noncovered services. Private health insurance
includes EBI plans and nongroup plans and excludes
medigap plans, Medicare Advantage plans, and Medicaid managed care plans. The
first of the two models, the current-year model, projects expenditures
for the first year of the forecast window using recent data on premiums
from surveys. The second, the primary model, projects expenditures for
the remainder of the forecast window using an autoregressive term and
CBO's macroeconomic forecast. The 2025 PGM uses growth in real income to
project real health expenditures per privately insured person. The
projected real health expenditures are then combined with projected
prices (PCEMED) to produce nominal projected expenditures per capita.

### Data

The key source for historical estimates of expenditures per capita is
the National Health Expenditure Accounts (NHEA).[^1] CBO modifies those
data in two ways. First, estimated medigap plan expenditures and
medigap enrollment are removed. Second, expenditures are adjusted to
remove the effects of onetime events that significantly affected
expenditure growth but should not be reflected in the long-term
equilibrium--namely the establishment of health insurance marketplaces
under the Affordable Care Act[^2] and the COVID-19 pandemic.

[^1]: For definitions of terms used in this report, please refer to the Glossary at the end of this report.

[^2]: The growth rate is increased by half a percentage point in 2013
    and decreased by half a percentage point in 2014 to account for
    the effects of the Affordable Care Act.

The ACA marketplaces significantly altered the individual insurance
market by expanding coverage and introducing new subsidies, but their
impact on expenditure growth rates varied over time. In the initial
years, there was a surge in enrollment, particularly among previously
uninsured individuals with preexisting health care needs, leading to a
temporary spike in expenditures. Over time, as the market stabilized and
risk pools adjusted, expenditure growth rates may have normalized, which
is why CBO views the establishment of those marketplaces more as a
onetime shock than a permanent change in trend.

CBO creates two versions of the historical and projected expenditures.
One version includes an adjustment to exclude any growth due to changes
in the age and sex composition of the privately insured population, and
the other version does not include that adjustment. HISIM2 uses premium
growth projections that exclude the effects of demographic changes
because the microsimulation model implicitly accounts for those changes
while projecting insurance coverage.[^3] To create the age- and
sex-adjusted version, CBO uses the Current Population Survey (CPS) to
estimate, for each historical year, the share of the privately insured
population with each combination of single year of age and sex. CBO then
applies the spending patterns estimated by
Dale H. Yamamoto (2013) to estimate the change in expenditures attributable to demographic
changes and creates an annual index used to scale estimates of health
expenditures per privately insured person.[^4]

[^3]: For the Joint Committee on Taxation, CBO produces a version of the
    projections that does not account for demographic changes. For that
    purpose, the series includes all other modifications discussed in
    this report and is projected using the same model specification.

[^4]: Dale H. Yamamoto. [*Health Care Costs---From Birth to Death*](https://www.soa.org/resources/research-reports/2013/research-health-care-birth-death/) (2013).

Other inputs to the model include [CBO's historical data and economic projections](https://www.cbo.gov/data/budget-economic-data#11).
Those variables include the total U.S. population (including members of the
armed forces overseas and the institutionalized population), personal
disposable income (PDI), the personal consumption expenditures price
index for consumer goods (PCE). For the PGM, CBO also uses historical and projected values of the
personal consumption expenditures price index for medical spending
(PCEMED).

CBO adjusts both the expenditure and income variables during the
pandemic period because the pandemic resulted in higher PDI and a
temporary decrease in utilization of health care services. That decrease
in utilization resulted in low expenditure growth in 2020 and high
expenditure growth in 2021 as utilization rebounded. Without adjustment,
those outliers may result in less accurate coefficient estimates and
affect the projections through the autoregressive terms in the model.
CBO adjusts income because stimulus payments in 2021 and 2022 provided a
temporary boost to income that was slowly spent down over subsequent
years. Less significantly, many people realized large capital gains in
2021 because of excess liquidity from the 2020 stimulus checks and
other pandemic-related fiscal measures. CBO smoothed those fluctuations
in income under the assumption that health spending responds more to
permanent changes in income than temporary ones over the medium run.

For the spring 2025 baseline, CBO made the following changes to its
treatment of the data:

| Factor | Spring 2024 | Spring 2025 |
|----|----|----|
| COVID Adjustment: Expenditures | Fixed expenditure growth to 2.4% (2020) and 2.5% (2021) to align with survey estimates of premium growth. | Fixed 2020 expenditures at the midpoint of 2019 and 2021 expenditures. |
| COVID Adjustment: PDI | None | Smoothed PDI to account for the slow draw down of stimulus income in 2021 and 2022 and surplus capital gains in 2021. |
| medigap backcasting[^5] | Backcast medigap spending to 1987 using Consumer Price Index for All Urban Consumers: Medical Care | Backcast medigap spending to 1987 using Medicare expenditure per capita |

[^5]: The NHEA estimates of medigap spending per enrollee begin in 2001.

### Methods

The PGM consists of two models: the current-year model and the primary
model. The current year model is used to project real health
expenditures per capita in the first year of the forecast window. The
primary model is used to project health expenditures for the remaining
years of the window. As the primary model is an autoregressive model,
the projection of the current-year model is an input to the projection
of the primary model.

The current-year regression model calculates a preliminary estimate of
growth in private health expenditures per capita from 2023 to 2024. That
model is estimated using historical data through 2023 on growth in
premiums for EBI plans and then used to project growth in
expenditures in 2024. This model is necessary because the last year of
historical data from the NHEA is 2023, but other sources provide useful
information on growth in private premiums from 2023 to 2024 and from 2024 to
2025. The sources for this regression model are the Bureau of Labor
Statistics' (BLS's)
producer price index (PPI) for comprehensive medical insurance plans, and
KFF's Employer Health Benefits
Survey (EHBS).[^6]

[^6]: See Bureau of Labor Statistics, [*PPI industry data for Direct health and medical insurance
    carriers-Comprehensive medical service plans, not seasonally
    adjusted*](https://data.bls.gov/timeseries/PCU524114524114101) (accessed January, 2025). For more
    information on the EHBS, see Gary Claxton, Matthew Rae, Aubrey Winger,
    and Emma Wager, [*Employer Health Benefits: 2024 Annual Survey*](https://files.kff.org/attachment/Employer-Health-Benefits-Survey-2024-Annual-Survey.pdf) (October 2024).
    

For 2025 and beyond, health expenditures per capita are projected using
the primary model. The primary model is an autoregressive model of
expenditure growth that includes projections of PDI and, new for the
spring 2025 baseline, medical prices (PCEMED).[^7] Thus, the model tends
to revert from expenditure growth in the first year of the window to a
long-term equilibrium while tracking projected changes in income and
medical prices. Although the historical NHEA data extend into the 1980s,
CBO excludes data prior to 1999 when estimating the model, as expenditure
growth in the 1990s was driven by the rise of and subsequent backlash
against health maintenance organizations (HMOs). CBO believes that the more
distant historical period is less applicable to projecting premium
growth over the next decade.

[^7]: For convenience, the model includes the PCEMED index as an
    explanatory variable with a fixed coefficient of one. That approach
    is equivalent to estimating the model and generating predictions in
    real terms then multiplying the predictions by PCEMED to convert
    them to nominal terms.

To further refine its estimates, CBO conducted a series of stakeholder
interviews and reviewed data from other sources including the Federal
Employees Health Benefits program. Stakeholders were selected to
represent the views of the insurance industry, actuaries, and employers
in their role as purchasers of health benefits. Drawing from those
sources, CBO applies an adjustment factor to the current-year estimate
(growth from 2023 to 2024), and the first year of the projection (growth
from 2024 to 2025).

The following table summarizes updates to the PGM methodology applied to
the spring 2025 baseline:

|  |  |  |
|----|----|----|
| **Factor** | Spring 2024 | Spring 2025 |
| Near-term projection | 2.0% growth from 2021 to 2022 and 6.7% growth from 2022 to 2023, based on review of NHEA data and survey sources | Preliminary estimate of growth from 2023 to 2024 based on current-year model; growth from 2023 to 2024 increased by 2 percentage points and growth from 2024 to 2025 increased by 2 percentage points on the basis of stakeholder interviews and other data |
| PGM: Estimation period | 2003-2022 (*N* = 20) | 1999-2023 (*N* = 25) |
| PGM: Dependent variable | Percentage change in nominal private health expenditures per capita | Log difference in health expenditures per capita deflated by PCEMED |
| PGM: Lagged dependent variable terms | Three | One |
| PGM: Income variable | Three-year log difference in nominal PDI per capita | Six-year log difference in PDI per capita deflated by PCE |
| PGM: Estimator | Prais-Winsten | Ordinary least squares |

## Changes to the Projection

To facilitate the comparison of the spring 2024 and spring 2025 projections,
changes in the estimates are decomposed into a series of steps:

1.  **Updating the Historical Data:** Changes resulting from updating the
    data on expenditures per capita (and our adjustments to them) and
    the demographic composition of the privately insured population. The
    macroeconomic forecast and the model specification and coefficients
    are held fixed.
2.  **Updating the Macroeconomic Forecast:** Changes resulting from updating
    the macroeconomic forecast. The model specification and coefficients
    are held fixed.
3.  **Updating the Coefficients of the Previous Model Specification:**
    Changes resulting from updating the coefficient estimates using the
    model specification from the previous baseline. The model
    specification is held fixed.
4.  **Updating the Model Specification:** Changes resulting from updating
    the model specification.
5.  **Adjusting the Near-Term Projections:** Adjustments are made to the
    current-year estimate and the estimate for the first year of the
    projection. These adjustments reflect stakeholder input and data not
    otherwise formally incorporated in the projection.

### Step 1: Updating the Historical Data

```{r calc_increase_step1, warning=FALSE, message=FALSE, results='hide'}
# Calculate the increase in private health expenditures per capita in 2034 
# between prem_demo_b2024 and prem_demo_b2025
increase <- pgm_output |>
    select(series, cy2034) |>
    filter(series %in% c("prem_demo_b2024", "prem_demo_b2024_step1")) |>
    spread(series, cy2034) |>
    mutate(
        pct_increase = (prem_demo_b2024_step1 - prem_demo_b2024) / prem_demo_b2024,
        dollar_increase = prem_demo_b2024_step1 - prem_demo_b2024
      )

pct_increase_step1 <- increase |>
    pull(pct_increase) |>
    # Format as a percentage
    scales::percent(accuracy = 0.1)

dollar_increase_step1 <- scales::label_dollar(accuracy=1)(increase |> pull(dollar_increase))
```

As shown in the chart below, CBO's projection of expenditures per capita
in 2034 has increased by `r dollar_increase_step1`
(`r pct_increase_step1`) since last year because of changes in the historical
data. In the historical data, values for expenditure growth increased in
2020, 2022, and 2023. Those changes yield higher premium growth in 2024
but a similar path thereafter. The difference in 2020 is due to the
aforementioned change in CBO's method for adjusting expenditure growth
during the first year of the pandemic, when expenditures fell sharply
relative to premiums. The difference in 2022 and 2023 reflects (1) an
upward revision to the NHEA data for 2022 and (2) a new data point in
2023. Both the revised 2022 and new 2023 premium growth rates are higher
than the rates estimated from survey data in those years.

```{r step1, warning=FALSE, message=FALSE, results='hide'}

# Step 1: Updating Historical NHEA and CPS Data
level_series_step1 <- c(level_prev, level_step1)  
pchange_series_step1 <- c(pchange_prev, pchange_step1)  
combined_plot_step1 <- forecast_plots(level_series_step1, pchange_series_step1, 
                                      "Step 1: Updating Historical NHEA and CPS Data")
combined_plot_step1
```

### Step 2: Updating the Macroeconomic Forecast

The updates to CBO's PDI forecast had a negligible effect on the
projections of health expenditure growth. This step includes the
adjustments to the PDI forecast in the early 2020s to account for
drawn-out spending of stimulus payments received in 2021 and 2022 and
excess capital gains in 2021. The effect of updating the PDI forecast is
minor because the estimated coefficient on the three-year growth rate of
PDI per capita in the spring 2024 baseline model was very small: A 10
percentage-point increase in the three-year PDI per capita growth rate
yields only a 0.2 percentage-point increase in the health expenditures
per capita growth rate. The change in PDI was negligible, as the
previous specification placed less emphasis on it.

```{r step2, warning=FALSE, message=FALSE, results='hide'}

# Step 2: Updating the Macroeconomic Forecast
level_series_step2 <- c(level_step1, level_step2)
pchange_series_step2 <- c(pchange_step1, pchange_step2)
combined_plot_step2 <- forecast_plots(level_series_step2, pchange_series_step2,
                                      "Step 2: Updating the Macroeconomic Forecast")
combined_plot_step2
```

### Step 3: Updating the Coefficients of the Previous Model Specification

In Step 3, CBO reestimates the spring 2024 specification using the updated
input data and projects expenditure growth using the updated coefficient
estimates. In the updated estimates, the effect of PDI roughly doubled,
and the magnitude of the coefficients on the first and second
autoregressive terms (which are positive and negative, respectively)
increased. Those changes yield a projected peak in expenditure growth in
2024, greater oscillation around the equilibrium thereafter, and,
overall, higher expenditures in 2033.

```{r step3, warning=FALSE, message=FALSE, results='hide'}

# Step 3: Updating Coefficients for the Previous Model Specification
level_series_step3 <- c(level_step2, level_step3)
pchange_series_step3 <- c(pchange_step2, pchange_step3)
combined_plot_step3 <- forecast_plots(level_series_step3, pchange_series_step3,
                                      "Step 3: Updating Coefficients for\nthe Previous Model Specification")
combined_plot_step3

```

### Step 4: Updating the Model Specification

The next step in updating the projection is estimating the new model
specification. One update is to define a formal regression model (the
"current-year" model) for growth in the first year of the projection (from 2023 to 2024). That model is a simple regression of expenditure
growth on survey estimates of premium growth from BLS and KFF's EHBS.
The PGM specification for 2025 was also updated in several key ways.

First, the projection is now based on real health expenditures per
capita, rather than nominal health expenditures per capita, and the projections of real health
expenditures are combined with a projected price index (PCEMED) to yield
projected nominal values. Second, the regression now includes real PDI
as a predictor, rather than nominal PDI. Those two changes reflect a
conceptual framework in which growth in real income drives growth in
real health expenditures, and price growth is modeled separately.

Third, growth in PDI is now measured over six rather than three years.
That allows for longer delays in the process by which real income leads
to changes in real health expenditures. CBO uses a six-year lag because
a literature review by the agency identified income as a primary
determinant of health spending growth in the medium term and model
validation statistics indicated that long-term growth rates are a better
predictor of health spending than short-term growth rates or distributed
lag models.

Fourth, the PGM now uses one rather than three autoregressive terms.
During model validation, CBO found that one lag was sufficient to
eliminate autocorrelation in the model residuals.

The change in model specification yields a sharply lower projection for
2025 and a smoother projection for the remainder of the window that ends
at a slightly lower rate of growth. According to CBO's conversations
with stakeholders, the previous specification's forecast for 2025 is
more plausible than the new specification's; however, it also yields an
implausible oscillating pattern in the remainder of the forecast window.
By using a specification with a single lagged term while imposing a
positive adjustment to the forecast in 2024 and 2025, the agency
achieves a similarly plausible projection for 2025 with a more plausible
path thereafter. Essentially, CBO's updated specification models growth
in 2024 and 2025 as the product of temporary shocks to omitted variables
in those years rather than as the product of a more complicated
autoregressive process.

```{r step4, warning=FALSE, message=FALSE, results='hide'}

# Step 4: Updating the Model Specification
level_series_step4 <- c(level_step3, level_step4)
pchange_series_step4 <- c(pchange_step3, pchange_step4)
combined_plot_step4 <- forecast_plots(level_series_step4, pchange_series_step4,
                                     "Step 4: Updating the Model Specification")
combined_plot_step4

```

### Step 5: Adjusting the Near-Term Projections

The final projections incorporate a 2 percentage-point boost in both 2024 and
2025. Those boosts were incorporated to
capture the findings from stakeholder interviews and other data
indicating rapid growth in expenditures over the near term. This initial
jump is expected to slow as a result of stabilization of take-up of high-cost
medication among enrollees in private health insurance.

```{r step5, warning=FALSE, message=FALSE, results='hide'}
level_series_step5 <- c(level_step4, level_current)
pchange_series_step5 <- c(pchange_step4, pchange_current)
combined_plot_step5 <- forecast_plots(level_series_step5, pchange_series_step5,
                                     "Step 4: Adjusting the Near-Term Projections")
combined_plot_step5
```

### Additional Projections: Health Care Expenditures Per Capita Without Demographic Adjustment

CBO also produces projections of health expenditures per capita that
incorporate changes in the age and sex composition of the privately
insured population. The agency generates those predictions by estimating
the same current-year and primary models using data that have not been
adjusted for demographics. The difference between the adjusted and
unadjusted projections (in log differences) represents the implied
change in health expenditures per capita due to demographic changes. As
growth in the unadjusted projections exceeds growth in the adjusted
projections, CBO is implicitly forecasting that demographic changes will
contribute positively to growth in health expenditures per capita over
the next 10 years. That expected trend is consistent with the continued
aging of the U.S. population.

```{r jct, warning=FALSE, message=FALSE, results='hide'}

level_series_jct <- c(level_jct_prev, level_jct_current)
pchange_series_jct <- c(pchange_jct_prev, pchange_jct_current)
combined_plot_jct <- forecast_plots(level_series_jct, pchange_series_jct,
                                      "Previous and Current Projections\nWithout Demographic Adjustments")
combined_plot_jct
```

\section*{Glossary}
\label{sec:glossary}

## Glossary

-   **Health expenditures per capita**

The premium growth model uses data on private enrollment and
expenditures from Table 21 of the National Health Expenditure Accounts
to create private per capita premiums. National health expenditures
equal health consumption expenditures plus the sum of medical-sector
purchases of structures and equipment and expenditures for noncommercial
medical research (investment).

Private Per Capita Premiums = Total Private Health Insurance
Expenditures / Total Private Health Insurance Enrollment

Sources: [Centers for Medicare & Medicaid Services, "Historical"](https://www.cms.gov/data-research/statistics-trends-and-reports/national-health-expenditure-data/historical)
(accessed January, 2025), and [Bureau of
Economic Analysis, NIPA Handbook: Concepts and Methods of the U.S.
National Income and Product Accounts, Chapter 5](https://www.bea.gov/resources/methodologies/nipa-handbook) (December 2024),
.

-   **Personal consumption expenditures price index for medical spending
    (PCEMED)**

The premium growth model uses PCEMED to measure medical prices and
deflate nominal private health insurance expenditures per capita. Historical values of PCEMED are taken from the Bureau of
Economic Analysis's (BEA's) National Income and Products Accounts
(NIPAs), and projections are done by CBO.

PCEMED has three subcomponents:

-   PCE: Therapeutic Appliances & Equipment Price Index , also known as PCDMED, which comprises 2 percent of PCEMED

-   PCE: Pharmaceutical & Other Medical Products Price Index, also known as PCNMED, which comprises 17 percent of PCEMED

-   PCE: Health Care Services, also known as PCSMED, which comprises 81 percent of PCMED

CBO's projections of PCDMED and PCNMED are based on simple
autoregressive moving average models. The agency's projections of
PCSMED, which makes up the bulk of the PCEMED index, involve an
iterative process within CBO between MAD and the Budget Analysis Division.

Key steps for projecting PCSMED include:

1.  Forecasting producer price indexes (PPIs) for inpatient care,
    outpatient care, and physician care

2.  Using those PPIs to forecast PCE price indexes for hospital care and
    physician care

3.  Aggregating the resulting PCE indexes along with the employment cost
    index (ECI) to forecast the overall PCE for medical services

Equation for PCSMED (as of spring 2024), where the Greek letter pi represents growth in a price index, beta represents an estimated coefficient, and epsilon is an error term:

$$ \pi_t^{\text{PCSMED}} = \beta_1\pi_{t-1}^{\text{PCSMED}} + \beta_2\pi_t^{\text{PCE for hospital care}} + \beta_3\pi_t^{\text{PCE for physician care}} + \beta_4\text{ECI}_t + \epsilon_t$$

The following table describes some of the key differences between PCEMED and the Consumer Price Index: Medical care (CPI-M).

| Factor | PCEMED | CPI-M |
|----|----|----|
| Scope | Includes direct payments made by consumers and third-party payments made on behalf of consumers | Only includes out-of-pocket expenditures made by consumers |
| Construction | Fisher price index | Laspeyres index |
| Data sourcing | Sector-specific PPIs and sector-specific CPIs | Consumer surveys |
| Usage as medical care deflator | Adjusting for purchasing power changes on personal consumption expenditures | Adjusting for purchasing power changes in out-of-pocket expenditures |

-   **Personal consumption expenditures price index for consumer goods
    (PCE)**

The premium growth model uses PCE to deflate growth in personal
disposable income. This series is taken and processed from the Bureau of
Economic Analysis's National Income and Products Accounts (NIPA) data.
In the NIPAs, final consumption expenditures by households or nonprofit
institutions serving households represent the portion of PCE for
services provided to households without explicit charge, like the value
of nonprofit college education exceeding tuition and fees. It equals
their gross output, calculated as current operating expenses (excluding
capital investments) minus sales to households and other sectors and the
value of investment goods.

Source: Congressional Budget Office, [*Historical Data and Economic Projections*](https://www.cbo.gov/data/budget-economic-data#11)
(accessed January, 2025).

-   **Personal Disposable Income (PDI)**

The PGM uses real PDI, meaning nominal PDI divided by PCE, as an
explanatory variable for real premium growth. PDI is the income that is
left after people pay their taxes, and so is also known as after-tax
income.

Source: Congressional Budget Office, [*Historical Data and Economic Projections*](https://www.cbo.gov/data/budget-economic-data#11)
(accessed January, 2025); and Bureau of Economic Analysis,
[*Income & Saving*](https://www.bea.gov/resources/learning-center/what-to-know-income-saving) (accessed January, 2025).

-   **Yamamoto index**

Dale H. Yamamoto measured health care spending among a sample of enrollees
in commercial health insurance, by single year of age and sex. Those
spending measures are combined with enrollment data from the Current
Population Survey and are used to estimate the role of demographic
changes in the historical trends in private health expenditures per
capita.

Source: Dale H. Yamamoto. [*Health Care Costs---From Birth to Death*](https://www.soa.org/resources/research-reports/2013/research-health-care-birth-death/) (2013).
